Live Functional Programming with Typed Holes

### 2018-04-11 related-work.tex
draft of sec 3.2 through 3.2.1 and a stub starting to talk about cast insertion
finished draft of sec3 and 4
uncommitted Related Work wip from several weeks ago
- breaking up the big list into smaller sections/lists
- importing some text from the grant

\begin{itemize}
	\item Hazelnut paper
	\item simply typed underdeterminism paper
	\item work on partial evaluation and staging (incl. connections to modal logic -- ``modal analysis of staged computation'')
	\item symbolic evaluation
	\item full beta reduction
	\item Agda
	\item Idris
	\item GHC holes
	\item CMTT
	\item gradual typing (and dynamic typing)
		\begin{itemize}
			\item siek and taha paper
			\item snapl15 paper
			\item gradualizer paper
			\item maybe other things, e.g. several papers by ron garcia
		\end{itemize}
	\item DuctileJ stuff -- \url{https://homes.cs.washington.edu/~mernst/pubs/ductile-icse2011.pdf}
	\item OLEG from McBride's thesis
	\item Visual Studio (and others) support for edit-and-resume
	\item Scratch lets you just skip over statement holes
	\item prior work on confluence for the lambda calculus
	\item work on debuggers that allow you to inspect environments
		\begin{itemize}
		\item might be something relevant in the paper ``A Debugger for Standard ML'' 
		\item "Visualizing the evaluation of functional programs for debugging" by Whitington and Ridge
		\item "A lightweight interactive debugger for Haskell'' and ``Multiple-View Tracing for Haskell: a New Hat'' might be relevant
		\item ocamli -- \url{https://github.com/johnwhitington/ocamli}
		\item Better supporting debugging aids learning a novel programming language. -- Scaffidi at VLHCC 2017
		\item quote from Wadler in ``Why no one uses functional languages'':
			\begin{quote}
			“...there are few debuggers or
profilers for strict [functional] languages, perhaps because constructing them is not considered
research. This is a shame, since such tools are sorely needed, and there remains much of
interest to learn about their construction and use.
\end{quote}
		\end{itemize}
	\item papers that show up in a search for ``typed holes'' -- \url{https://scholar.google.com/scholar?hl=en&as_sdt=0%2C39&q=%22typed+holes%22&btnG=}
	\item maybe also search for ``partial programs'' 
	\item roly pererra work on program slicing
	\item should say something about how holes show up in program synthesis under various names (what?) 
	\item ``Achieving flexibility in direct-manipulation programming environments by relaxing the edit-time grammar'' -- \url{http://ieeexplore.ieee.org/document/1509511/}
	\item ``Call-by-value is dual to call-by-name'' might be relevant? \url{http://homepages.inf.ed.ac.uk/wadler/papers/dual/dual.pdf}
	\item mention how unspecified evaluation order is something that people do when talking about parallelism?
\end{itemize}

>>>>

Program editors frequently encounter program text that is not well-formed with
respect to the concrete syntax of the language \cite{popl-paper}.
%
Within the simplest editors, editor services that take well-formed program text
or an abstract syntax tree as input are simply disabled when this occurs.
%
More sophisticated editors, including those generated by modern editor
generators like Spoofax \cite{DBLP:conf/oopsla/KatsV10}, deploy \emph{error
recovery} heuristics that silently insert tokens or discard tokens from the
program so that the editor-internal representation is a standard abstract syntax
tree
\cite{DBLP:journals/siamcomp/AhoP72,charles1991practical,graham1979practical,DBLP:conf/oopsla/KatsJNV09}.
%
These heuristics are typically defined manually by the grammar designer, though
certain heuristics can be generated semi-automatically by tools that are given a
description of the scoping conventions of the language and of secondary
notational conventions (\eg{}~whitespace conventions)
\cite{DBLP:conf/oopsla/KatsJNV09,DBLP:conf/sle/JongeNKV09}.
%
These error recovery heuristics require guessing at the programmer's intent, and
result in tokens being inserted or deleted from the program invisibly, so they
are \emph{ad hoc} and can confuse the programmer
\cite{DBLP:conf/oopsla/KatsJNV09}.

\parahead{Hole-Driven Development}

A cleaner alternative approach, and the approach that is central to our proposed
research, is to add \emph{holes} to both the concrete and abstract syntax of the
language.
%
Syntax trees with holes first arose in prior work on \emph{structure editors},
which are program editors where every edit action directly causes changes to the
abstract syntax tree of the program, rather than going indirectly through a
parser.
%
Structure editors have a long history: the Cornell Program Synthesizer was
developed in the early 1980s \cite{teitelbaum_cornell_1981} and there has been a
steady stream of new designs in this space
(\eg{}~\cite{ko_barista:_2006,Resnick:2009:SP:1592761.1592779,voelter_mbeddr:_2012,tillmann_touchdevelop:_2011,lamdu,ActiveCodeCompletion,Greenfoot,Deuce}).
%
For example, Scratch and other block-based editors have achieved success as a
tool for teaching children how to program
\cite{Resnick:2009:SP:1592761.1592779}. \texttt{mbeddr} is an editor for a
C-like language \cite{voelter_mbeddr:_2012}, built using the commercially
supported MPS structure editor workbench \cite{voelter2011language}, that is
presently deployed to a number of customers.
%
TouchDevelop is a structure editor for an object-oriented language designed for
use on mobile devices \cite{tillmann_touchdevelop:_2011}. Lamdu \cite{lamdu},
Unison \cite{unison} and  isomorf~\cite{isomorf} are open source structure
editors for functional languages similar to ML.

%% There is reason to be optimistic, however, with recent studies 
%% suggesting that programmers experienced with a modern keyboard-driven structure editor like \texttt{mbeddr} 
%% can be highly productive \cite{DBLP:conf/vl/Asenov014,DBLP:conf/sle/VolterSBK14}.

Nevertheless, text-based program editors continue to dominate in practice, in
part because text is flexible and familiar and because tools for interacting
with text, like regular expressions and various differencing techniques used by
version control systems, are already very well-developed.
%
Consequently, researchers have also explored various ``hybrid'' approaches,
which incorporate holes into an otherwise text-based program editor.
%
For example, recent iterations of the Glasgow Haskell Compiler (GHC)
\cite{GHCWIKI}, as well as the dependently typed proof assistants
Idris~\cite{brady2013idris} and Agda~\cite{norell2009dependently}, support a
workflow where the programmer places holes manually at locations in the program
that remain under construction. Agda and Idris also allow the programmer to put
terms inside holes, to indicate that the term may not yet be well-typed or
otherwise finished.
%
Recent work on \emph{syntactic placeholders} takes some of the tedium out of
this approach by providing low-level edit actions that cause these holes (called
placeholders in their paper) to be inserted automatically \cite{Amorim2016}. 
%
This approximates the experience of a structure editor in common usage, while
allowing the programmer to perform arbitrary text edits when necessary.
%
Expression holes also appear in the context of program synthesis, serving as
placeholders in \emph{templates}~\cite{srivastava2013template} or
\emph{sketches}~\cite{solar2009sketching} to be filled in by an expression
synthesis engine.

\begin{itemize}
	%% \item Agda
	%% \item Idris
	%% \item GHC holes
	%% \item Visual Studio (and others) support for edit-and-resume
	\item Scratch lets you just skip over statement holes
	\item papers that show up in a search for ``typed holes'' -- \url{https://scholar.google.com/scholar?hl=en&as_sdt=0%2C39&q=%22typed+holes%22&btnG=}
\end{itemize}

\parahead{Structure Editor Calculus}

Most work on structure editors has focused on the user interfaces that they
present. This is important work---presenting a fluid user interface involving
exclusively structural edit actions is a non-trivial problem that has not yet
been fully resolved, though recent studies have started to show productivity
gains for keyboard-driven structure
editors~\cite{DBLP:conf/vl/Asenov014,DBLP:conf/sle/VolterSBK14}.

We emphasize
that our proposed contributions will be relevant no matter how holes come to be
inserted into the program. % We will say more about all of these systems later.

two main remaining questions (beyond scaling up the calculus) are how to run
programs with type and expression holes.

\begin{itemize}
	\item Hazelnut paper
\end{itemize}

\parahead{Gradual Typing}

As already discussed in Track 1, type holes are closely related to the unknown
types of gradual type theory \cite{Siek06a}. This has implications for our
dynamic semantics as well---we need to be able to run programs where type holes
have obscured a type inconsistency, \eg{}~\li{let f(g : }\ \li{) => g 3 in f 4}.
This term is well-typed according to the static semantics of Track 1 because the
argument \li{g} of \li{f} has an unknown type, and so it can be applied to
\li{3}, but when \li{f} is applied to \li{4}, there is a problem because \li{4 3}
is meaningless. There are two ways to address the problem. The first is to
apply ML-style type inference to discover a more specific type schema for
\li{g}, \ie{}~that it must actually be an arrow type, and then mark the argument
\li{4} to \li{f} as a type inconsistency using a non-empty hole. We will
investigate this in Track 1. The second approach is to adopt run-time cast
insertions from research on gradual type theory. We will investigate this second
approach in Track 2.

\begin{itemize}
	\item gradual typing (and dynamic typing)
  \item siek and taha paper
  \item snapl15 paper
  \item gradualizer paper
  \item maybe other things, e.g. several papers by ron garcia
\end{itemize}

\parahead{Contextual Modal Type Theory}

compared to the dynamics of type holes (which are based on Grad Typing), the
dynamics of expression holes are less clear.

Formally, expression holes in our proposed dynamics are closely related to the
metavariables of contextual modal type theory (CMTT) \cite{Nanevski2008}.
Metavariables in CMTT are associated with both a type and a typing context.
Expression holes can similarly be assigned a most specific type and typing
context. The environment tracking service can then be understood as being
related to the concept of a \emph{metavariable closure} in CMTT, which consists
of a metavariable paired with a substitution for all of the variables in the
typing context associated with that metavariable. Indeed, this neatly explains
the difference between our proposed dynamics and existing work on staging and
partial evaluation---that existing work is rooted in modal type theory
\cite{Davies:2001op} rather than the derivative contextual modal type theory
(which, to date, has been studied far less extensively.) In other words,
existing systems are focused on partial evaluation with respect to an input to a
function, whereas holes can appear in context, \eg{}~directly within the body of
a function. Recent work on the Beluga theorem prover is also based on CMTT,
albeit for the purpose of proof refinement and contextual reasoning rather than
program evaluation \cite{DBLP:conf/flops/Pientka10,pientka2015inductive}.
McBride's OLEG is another system for reasoning contextually about metavariables
\cite{DBLP:phd/ethos/McBride00}, and it is the basis of certain analagous
features of Idris \cite{brady2013idris}.

\begin{itemize}
	\item CMTT
\end{itemize}

\parahead{Evaluation Order}

\begin{itemize}
	\item full beta reduction
	\item prior work on confluence for the lambda calculus
	\item ``Call-by-value is dual to call-by-name'' might be relevant? \url{http://homepages.inf.ed.ac.uk/wadler/papers/dual/dual.pdf}
	\item mention how unspecified evaluation order is something that people do when talking about parallelism?
\end{itemize}

\parahead{Partial Evaluation and Symbolic Execution}

\begin{itemize}
	\item work on partial evaluation and staging (incl. connections to modal logic -- ``modal analysis of staged computation'')
	\item symbolic evaluation
	\item maybe also search for ``partial programs'' 
\end{itemize}

put Seidel here

\parahead{Debugging}

\begin{itemize}
	\item work on debuggers that allow you to inspect environments
  \item might be something relevant in the paper ``A Debugger for Standard ML'' 
  \item "Visualizing the evaluation of functional programs for debugging" by Whitington and Ridge
  \item "A lightweight interactive debugger for Haskell'' and ``Multiple-View Tracing for Haskell: a New Hat'' might be relevant
  \item ocamli -- \url{https://github.com/johnwhitington/ocamli}
  \item Better supporting debugging aids learning a novel programming language. -- Scaffidi at VLHCC 2017
  \item quote from Wadler in ``Why no one uses functional languages'':
    \begin{quote}
    “...there are few debuggers or
profilers for strict [functional] languages, perhaps because constructing them is not considered
research. This is a shame, since such tools are sorely needed, and there remains much of
interest to learn about their construction and use.
    \end{quote}
   \item edit-and-resume in visual studio, others
\end{itemize}

Visual Studio edit-and-continue \url{https://docs.microsoft.com/en-us/visualstudio/debugger/edit-and-continue}

Hole-like constructs also appear in work on program slicing
\cite{DBLP:conf/icfp/PereraACL12}, where holes are \rkc{dynamically inserted...}

\parahead{Others?}

\begin{itemize}
	\item simply typed underdeterminism paper
	\item DuctileJ stuff -- \url{https://homes.cs.washington.edu/~mernst/pubs/ductile-icse2011.pdf}
	\item ``Achieving flexibility in direct-manipulation programming environments by relaxing the edit-time grammar'' -- \url{http://ieeexplore.ieee.org/document/1509511/}
	\item talk about closures in environmental semantics
\end{itemize}
